---
title: "Correlations Part I"
author: "Lauren Blake"
date: 2018-07-18
output: workflowr::wflow_html
---




# Introduction

The goal of this script is to look for correlations in interindividual features.  

# Individual-specific measures

```{r}
# Load library for plotting
library(cowplot)
library(ggplot2)

# Load data
ind_only <- read.csv("../data/notimecovariates.csv")
str(ind_only)
```

# Look at all correlations

```{r}
ind_only_characteristics <- c(2,3,4,7,8,9,10,11,12,13)
check_cor <- ind_only[,ind_only_characteristics]

# Obtain correlations and pvalues

pvalues <- matrix(data = NA, nrow = ncol(check_cor), ncol = ncol(check_cor))
colnames(pvalues) <- colnames(check_cor)
rownames(pvalues) <- colnames(check_cor)
correlations <- matrix(data = NA, nrow = ncol(check_cor), ncol = ncol(check_cor))
colnames(correlations) <- colnames(check_cor)
rownames(correlations) <- colnames(check_cor)

j=1
for (j in 1:ncol(check_cor)){
  
  for (i in 1:ncol(check_cor)){
      
  test <- cor.test(check_cor[,j], check_cor[,i], method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

correlations
pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05
```

Significant correlations: 

* EDU_days and bmiT1T2
* EDU_days and bmiT2T3
* bmiT1T2 and bmiT2T3
* bmiT2T3 and bmiT3T4
* EDU_days and blooddrawT1T2
* bmiT1T2 and blooddrawT1T2
* bmiT2T3 and blooddrawT1T2
* AMA and bmiT2T3
* age_onset and bmiT3T4

Suprisingly, AMA and EDU_days/AMA and days between T1 and T2 blood draws were not significantly correlated (maybe because those indviduals staying less than one week were not counted)?

# Plot of AMA and EDU day

```{r}
AMA_EDU <- ggplot(ind_only, aes(as.factor(ind_only$AMA), ind_only$EDU_days)) + geom_boxplot() + geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) + xlab("AMA status") + ylab("EDU days") + ggtitle("EDU days versus AMA status")

plot_grid(AMA_EDU)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/AMA_EDU.png", AMA_EDU,
#          base_aspect_ratio = 1)

```


# Plots of significant correlations

## EDU_days and bmiT1T2
```{r}
EDU_change <- ggplot(ind_only, aes(ind_only$EDU_days, ind_only$bmiT1T2)) + geom_point() + xlab("EDU days") + ylab("BMI change (T1 to T2)") + ggtitle("BMI change versus EDU days")

plot_grid(EDU_change)

save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/EDU_change_bmiT1T2.png", EDU_change,
          base_aspect_ratio = 1)

```

## EDU_days and bmiT2T3

```{r}
EDU_change <- ggplot(ind_only, aes(ind_only$EDU_days, ind_only$bmiT2T3)) + geom_point() + xlab("EDU days") + ylab("BMI change (T2 to T3)") + ggtitle("BMI change versus EDU days")

plot_grid(EDU_change)

save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/EDU_change_bmiT2T3.png", EDU_change,
          base_aspect_ratio = 1)

```

## bmiT1T2 and bmiT2T3

```{r}
EDU_change <- ggplot(ind_only, aes(ind_only$bmiT1T2, ind_only$bmiT2T3)) + geom_point() + xlab("BMI change (T1 to T2)") + ylab("BMI change (T2 to T3)") + ggtitle("BMI changes")

plot_grid(EDU_change)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/bmiT1T2_bmiT2T3.png", EDU_change,
#          base_aspect_ratio = 1)

```

## bmiT2T3 and bmiT3T4

```{r}
EDU_change <- ggplot(ind_only, aes(ind_only$bmiT2T3, ind_only$bmiT3T4)) + geom_point() + xlab("BMI change (T2 to T3)") + ylab("BMI change (T3 to T4)") + ggtitle("BMI changes")

plot_grid(EDU_change)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/bmiT2T3_bmiT3T4.png", EDU_change,
#          base_aspect_ratio = 1)

```

## AMA and bmiT2T3


```{r}
AMA_EDU <- ggplot(ind_only, aes(as.factor(ind_only$AMA), ind_only$bmiT2T3)) + geom_boxplot() + geom_dotplot(binaxis='y', stackdir='center', dotsize=0.5) + xlab("AMA status") + ylab("BMI change (T2 to T3)") + ggtitle("BMI change (T2 to T3) versus AMA status")

plot_grid(AMA_EDU)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/AMA_EDU.png", AMA_EDU,
#          base_aspect_ratio = 1)
```

## age_onset and bmiT3T4

```{r}
EDU_change <- ggplot(ind_only, aes(ind_only$age_onset, ind_only$bmiT3T4)) + geom_point() + xlab("Age of onset") + ylab("BMI change (T3 to T4)") + ggtitle("Age of onset versus BMI change (T3 to T4")

plot_grid(EDU_change)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/age_onset_bmiT3T4.png", EDU_change,
#          base_aspect_ratio = 1)

```

# Technical variables

```{r}
RIN_over_time <- read.csv("../data/RIN_over_time.csv")

RIN_by_time <- ggplot(RIN_over_time, aes(as.factor(RIN_over_time[,3]), RIN_over_time[,2])) + geom_boxplot(aes(fill=as.factor(RIN_over_time[,3]))) + geom_jitter(width = 0.15) +  xlab("Time") + ylab("RIN score") + ggtitle("RIN scores by timepoint (T1-T3)") + scale_fill_manual(name = "Time", values=c("#F8766D", "#B79F00", "#00BA38"))

plot_grid(RIN_by_time)

summary(lm(RIN_over_time[,2] ~ as.factor(t(RIN_over_time[,3]))))

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/RIN_by_time.png", RIN_by_time,
#          base_aspect_ratio = 1)
```

# On age, age of onset, and hospitalizations

```{r}
# Look at correlations

cor(ind_only$age_onset, ind_only$age, "pairwise.complete.obs")
cor(ind_only$Previous_hospitalizations, ind_only$age, "pairwise.complete.obs")
cor(ind_only$age_onset, ind_only$Previous_hospitalizations, "pairwise.complete.obs")

ages <- ggplot(ind_only, aes(ind_only$age_onset, ind_only$age)) + geom_point() + xlab("Age of onset") + ylab("Age at T1") + ggtitle("Age at T1 versus age of onset")

plot_grid(ages)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Age_age_onset.png", ages,
#          base_aspect_ratio = 1)

prev_hosp <- ggplot(ind_only, aes(ind_only$age_onset, ind_only$Previous_hospitalizations)) + geom_point() + xlab("Age of onset") + ylab("Previous hospitalizations") + ggtitle("Age of onset versus prev. hospital.")

plot_grid(prev_hosp)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Age_onset_prev_hosp.png", prev_hosp,
#          base_aspect_ratio = 1)
```




