---
title: "Correlations over time"
author: "Lauren Blake"
date: 2018-07-18
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---




# Introduction

The goal of this script is to look for correlations between clinical variables. 

```{r}
# Load library for plotting
library(cowplot)
library(ggplot2)
library(gplots)

# Load data
ind_only <- read.csv("/Users/laurenblake/Dropbox/Lauren Blake/RNA_seq_project/clinical_sample_info.csv")
str(ind_only)
dim(ind_only)

summary(ind_only)
```

# Weight over time

```{r}
weight_over_time <- ggplot(ind_only[1:144,], aes(as.factor(ind_only$time[1:144]), ind_only$Weight[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Time") + ylab("Weight (pounds)") + ggtitle("Weight changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/weight_over_time.png", weight_over_time,
#          base_aspect_ratio = 1)
```

# Weight over time (boxplot)

```{r}
weight_over_time <- ggplot(ind_only, aes(as.factor(ind_only$Time), ind_only$Weight)) + geom_boxplot(aes(color = as.factor(ind_only$Time)), outlier.shape = NA) + geom_jitter(aes(as.factor(ind_only$Time), ind_only$Weight), size = 0.5, width = 0.1) + theme(legend.position="none") + xlab("Timepoint") + ylab("Weight (lbs)") 

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Weight_over_time_boxplot.png", weight_over_time,
#          base_aspect_ratio = 1)
```


# BMI over time (boxplot)

```{r}
weight_over_time <- ggplot(ind_only, aes(as.factor(ind_only$Time), ind_only$bmi)) + geom_boxplot(aes(color = as.factor(ind_only$Time)), outlier.shape = NA) + geom_jitter(aes(as.factor(ind_only$Time), ind_only$bmi), size = 0.5, width = 0.1) + theme(legend.position="none") + xlab("Timepoint") + ylab("BMI") 

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/BMI_over_time_boxplot.png", weight_over_time,
#          base_aspect_ratio = 1)
```


# Weight over days (rather than T1, T2, etc.)

```{r}
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$Time_from_admission[1:144], ind_only$Weight[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Days from admission to blood draw") + ylab("Weight (pounds)") + ggtitle("Weight changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/weight_over_time_normalized.png", weight_over_time,
#          base_aspect_ratio = 1)
```

# Weight over days (rather than T1, T2, etc.)

```{r}
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$Time_from_admission[1:144], ind_only$Weight[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed") + geom_point(aes(color = as.factor(ind_only$current_ED[1:144]))) + xlab("Days from admission to blood draw") + ylab("Weight (pounds)") + ggtitle("Weight changes over time")  + labs(color='Eating disorder status') 

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/weight_over_time_by_recovery_status.png", weight_over_time,
#          base_aspect_ratio = 1)
```

# Look at all correlations for numeric variables

```{r}
# Eliminate factors
check_cor <- ind_only[,-1]
check_cor <- check_cor[,1:(ncol(check_cor)-6)]

# Obtain correlations and pvalues

pvalues <- matrix(data = NA, nrow = ncol(check_cor), ncol = ncol(check_cor))
colnames(pvalues) <- colnames(check_cor)
rownames(pvalues) <- colnames(check_cor)
correlations <- matrix(data = NA, nrow = ncol(check_cor), ncol = ncol(check_cor))
colnames(correlations) <- colnames(check_cor)
rownames(correlations) <- colnames(check_cor)

j=1
for (j in 1:ncol(check_cor)){
  
  for (i in 1:ncol(check_cor)){
      
  test <- cor.test(as.numeric(check_cor[,j]), as.numeric(check_cor[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05
```

# How highly correlated are the lab measurements?

```{r}
# Pull lab values
lab_values <- c(11:46, 50:53)
check_cor_labs <- ind_only[,lab_values]

# Obtain correlations and pvalues

pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))

heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none",
         notecex = 2,
          trace = "none", key = TRUE)




# Run on just the

lab_values <- c(7, 11:46, 50:53)
check_cor_labs <- ind_only[,lab_values]
lab_values_1 <- check_cor_labs[which(check_cor_labs$time == 1),2:40]
lab_values_2 <- check_cor_labs[which(check_cor_labs$time == 2),2:40]
lab_values_3 <- check_cor_labs[which(check_cor_labs$time == 3),2:40]

check_cor_labs <- lab_values_1


pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))


heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none",
         notecex = 2,
          trace = "none", key = TRUE)
```



# How do labs change over time? With changes in BMI? With age?

```{r}
# WBC versus time
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$Time_from_admission[1:144], ind_only$WBC[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Days from admission to blood draw") + ylab("WBC count") + ggtitle("WBC changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/WBC_over_time_normalized.png", weight_over_time,
#          base_aspect_ratio = 1)



# WBC versus BMI
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$bmi[1:144], ind_only$WBC[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("BMI") + ylab("WBC count") + ggtitle("WBC count versus BMI")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/WBC_versus_BMI.png", weight_over_time,
#          base_aspect_ratio = 1)

# WBC versus age
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$age[1:144], ind_only$WBC[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Age (years)") + ylab("WBC count") + ggtitle("WBC count versus age")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/WBC_versus_age.png", weight_over_time,
#          base_aspect_ratio = 1)
```

# How do labs change over time? With changes in BMI? With age?

```{r}
# ESR versus time
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$Time_from_admission[1:144], ind_only$ESR[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Days from admission to blood draw") + ylab("ESR") + ggtitle("ESR changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/ESR_over_time_normalized.png", weight_over_time,
#          base_aspect_ratio = 1)

# ESR versus BMI
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$bmi[1:144], ind_only$ESR[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("BMI") + ylab("ESR") + ggtitle("ESR versus BMI")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/ESR_versus_BMI.png", weight_over_time,
#          base_aspect_ratio = 1)

# ESR versus age
weight_over_time <- ggplot(ind_only[1:144,], aes(ind_only$age[1:144], ind_only$WBC[1:144], group = ind_only$BAN_ID[1:144])) + geom_line(linetype = "dashed", aes(color = as.factor(ind_only$BAN_ID[1:144]))) + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Age (years)") + ylab("ESR") + ggtitle("ESR count versus age")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/ESR_versus_age.png", weight_over_time,
#          base_aspect_ratio = 1)
```

# Labs over time

```{r}
weight_over_time <- ggplot(ind_only[1:144,], aes(as.factor(ind_only$Time[1:144]), ind_only$WBC[1:144])) + 
geom_boxplot() + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Time") + ylab("WBC count") + ggtitle("WBC changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/WBC_over_time.png", weight_over_time,
#          base_aspect_ratio = 1)



weight_over_time <- ggplot(ind_only[1:144,], aes(as.factor(ind_only$Time[1:144]), ind_only[1:144,12])) + 
geom_boxplot() + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Time") + ylab("RBC count") + ggtitle("RBC changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/RBC_over_time.png", weight_over_time,
#          base_aspect_ratio = 1)


weight_over_time <- ggplot(ind_only[1:144,], aes(as.factor(ind_only$Time[1:144]), ind_only[1:144,13])) + 
geom_boxplot() + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Time") + ylab("RBC count") + ggtitle("RBC changes over time")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/RBC_over_time.png", weight_over_time,
#          base_aspect_ratio = 1)


# define a function to remove outliers
#FindOutliers <- function(data) {
#  lowerq = quantile(data)[2]
#  upperq = quantile(data)[4]
#  iqr = upperq - lowerq #Or use IQR(data)
  # we identify extreme outliers
#  extreme.threshold.upper = (iqr * 3) + upperq
#  extreme.threshold.lower = lowerq - (iqr * 3)
#  result <- which(data > extreme.threshold.upper | data < extreme.threshold.lower)
# }

# use the function to identify outliers

#no_outliers <- check_cor_labs[which(complete.cases(check_cor_labs) == TRUE),]

#for (i in 1:ncol(no_outliers)){
#temp <- FindOutliers(no_outliers[,i])
#print(i)
#print(temp)
#}

weight_over_time <- ggplot(ind_only[1:144,], aes(as.factor(ind_only$Time[1:144]), ind_only$AST[1:144])) + 
geom_boxplot() + geom_point(aes(color = as.factor(ind_only$BAN_ID[1:144]))) + theme(legend.position="none") + xlab("Time") + ylab("AST") + ggtitle("Aspartate aminotransferase")

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/AST_over_time.png", weight_over_time,
#          base_aspect_ratio = 1)

```




