---
title: "voom_limma"
author: "Lauren Blake"
date: "2018-08-13"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

## Load raw counts and clinical data

```{r}
# Library

library(edgeR)
library(limma)
library(VennDiagram)
library(ggplot2)
library(cowplot)

# Read in the data

tx.salmon <- readRDS("../data/counts_hg38_gc_txsalmon.RData")
salmon_counts<- as.data.frame(tx.salmon$counts)


#tx.salmon <- readRDS("../data/counts_hg38_gc_dds.RData")
#salmon_counts<- as.data.frame(tx.salmon)


# Subset to T1-T3

salmon_counts <- salmon_counts[,1:144]

# Read in the clinical covariates

clinical_sample_info <- read.csv("../data/lm_covar_fixed_random.csv")
dim(clinical_sample_info)

# Subset to T1-T3

clinical_sample <- clinical_sample_info[1:144,(-12)]

dim(clinical_sample)
```

## Differential expression pipeline: TMM+voom+limma

```{r}
# Filter lowly expressed reads

cpm <- cpm(salmon_counts, log=TRUE)

expr_cutoff <- 1.5
hist(cpm, main = "log2(CPM) values in unfiltered data", breaks = 100, xlab = "log2(CPM) values")
abline(v = expr_cutoff, col = "red", lwd = 3)

hist(cpm, main = "log2(CPM) values in unfiltered data", breaks = 100, xlab = "log2(CPM) values", ylim = c(0, 100000))
abline(v = expr_cutoff, col = "red", lwd = 3)


# Basic filtering
cpm_filtered <- (rowSums(cpm > 1.5) > 72)

genes_in_cutoff <- cpm[cpm_filtered==TRUE,]

hist(as.numeric(unlist(genes_in_cutoff)), main = "log2(CPM) values in filtered data", breaks = 100, xlab = "log2(CPM) values")

# Find the original counts of all of the genes that fit the criteria 

counts_genes_in_cutoff <- salmon_counts[cpm_filtered==TRUE,]
dim(counts_genes_in_cutoff)

# Filter out hemoglobin

counts_genes_in_cutoff <-  counts_genes_in_cutoff[which( rownames(counts_genes_in_cutoff) != "HBB" ),]
counts_genes_in_cutoff <-  counts_genes_in_cutoff[which( rownames(counts_genes_in_cutoff) != "HBA2" ),]
counts_genes_in_cutoff <-  counts_genes_in_cutoff[which( rownames(counts_genes_in_cutoff) != "HBA1" ),]

# Take the TMM of the counts only for the genes that remain after filtering

dge_in_cutoff <- DGEList(counts=as.matrix(counts_genes_in_cutoff), genes=rownames(counts_genes_in_cutoff), group = as.character(t(clinical_sample$Individual)))
dge_in_cutoff <- calcNormFactors(dge_in_cutoff)

cpm_in_cutoff <- cpm(dge_in_cutoff, normalized.lib.sizes=TRUE, log=TRUE)

# Run PCA on the cpm data

pca_genes <- prcomp(t(cpm_in_cutoff), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
PC1 <- matrixpca[,1]
PC2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]

pcs <- data.frame(PC1, PC2, pc3, pc4, pc5)

summary <- summary(pca_genes)

head(summary$importance[2,1:5])

norm_count <- ggplot(data=pcs, aes(x=PC1, y=PC2, color= as.factor(clinical_sample$Time))) + geom_point(aes(colour = as.factor(clinical_sample$Time))) + ggtitle("PCA of normalized counts") + scale_color_discrete(name = "Time")

plot_grid(norm_count)

# Make sure certain clinical factors are factors (e.g. presence of psychiatric meds or not)

clinical_sample[,1] <- as.factor(clinical_sample[,1])

clinical_sample[,2] <- as.factor(clinical_sample[,2])

clinical_sample[,4] <- as.factor(clinical_sample[,4])

clinical_sample[,5] <- as.factor(clinical_sample[,5])

clinical_sample[,6] <- as.factor(clinical_sample[,6])


# Create the design matrix

# Use the standard treatment-contrasts parametrization. See Ch. 9 of limma
# User's Guide.

design <- model.matrix(~as.factor(Time) + as.factor(psychmeds) + RBC + RIN, data = clinical_sample)
colnames(design) <- c("Intercept", "Time2", "Time3", "Psychmeds", "RBC", "RIN")

# Fit model 

# Model individual as a random effect.
# Recommended to run both voom and duplicateCorrelation twice.
# https://support.bioconductor.org/p/59700/#67620

cpm.voom <- voom(dge_in_cutoff, design, normalize.method="none")
#check_rel <- duplicateCorrelation(cpm.voom, design, block = clinical_sample$Individual)
check_rel_correlation <-  0.1639035
cpm.voom.corfit <- voom(dge_in_cutoff, design, normalize.method="none", plot = TRUE, block = clinical_sample$Individual, correlation = check_rel_correlation)
#check_rel <- duplicateCorrelation(cpm.voom.corfit, design, block = clinical_sample$Individual)
check_rel_correlation <- 0.1638882

# Look at the densities of the gene expression data 

plotDensities(cpm.voom.corfit[,1])
plotDensities(cpm.voom.corfit[,2])
plotDensities(cpm.voom.corfit[,3])
           
# PCA on the filtered, normalized data

pca_genes <- prcomp(t(cpm.voom.corfit$E), scale = T, retx = TRUE, center = TRUE)

matrixpca <- pca_genes$x
PC1 <- matrixpca[,1]
PC2 <- matrixpca[,2]
pc3 <- matrixpca[,3]
pc4 <- matrixpca[,4]
pc5 <- matrixpca[,5]
pc6 <- matrixpca[,6]

pcs <- data.frame(PC1, PC2, pc3, pc4, pc5, pc6)

summary <- summary(pca_genes)

head(summary$importance[2,1:6])

# PCA by time

norm_count <- ggplot(data=pcs, aes(x=PC1, y=PC2, color=clinical_sample$Time)) + geom_point(aes(colour = as.factor(clinical_sample$Time)))  + scale_color_discrete(name = "Time") + xlab("PC1 (25.1% of variance)") + ylab("PC2 (13.0% of variance)") +  scale_colour_manual(name = "Time", values=c("#E69F00",  "#009E73", "#CC79A7", "#0072B2", "#D55E00"))

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Hg38_PC12_time12.png", norm_count,
#          base_aspect_ratio = 1)

norm_count <- ggplot(data=pcs, aes(x=pc3, y=pc4, color=clinical_sample$Time)) + geom_point(aes(colour = as.factor(clinical_sample$Time)))  + scale_color_discrete(name = "Time") + xlab("PC3 (8.8% of variance)") + ylab("PC4 (5.5% of variance)") +  scale_colour_manual(name = "Time", values=c("#E69F00",  "#009E73", "#CC79A7", "#0072B2", "#D55E00"))

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Hg38_PC34_time12.png", norm_count,
#          base_aspect_ratio = 1)

norm_count <- ggplot(data=pcs, aes(x=pc5, y=pc6, color=clinical_sample$Time)) + geom_point(aes(colour = as.factor(clinical_sample$Time)))  + scale_color_discrete(name = "Time") + xlab("PC5 (3.3% of variance)") + ylab("PC6 (2.5% of variance)") +  scale_colour_manual(name = "Time", values=c("#E69F00",  "#009E73", "#CC79A7", "#0072B2", "#D55E00"))

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/Hg38_PC56_time12.png", norm_count,
#          base_aspect_ratio = 1)



#write.csv(cpm.voom.corfit$E, #"../data/gene_expression_filtered_T1T3.csv")

#write.csv(pcs, "../data/gene_expression_filtered_pcs_T1T3.csv")

# Run lmFit and eBayes in limma

fit <- lmFit(cpm.voom.corfit, design, block=clinical_sample$Individual, correlation=check_rel_correlation)

# In the contrast matrix, have the time points

cm1 <- makeContrasts(Time1v2 = Time2, Time2v3 = Time3 - Time2, levels = design)

#cm1 <- makeContrasts(Time1v2 = Time2, Time2v3 = Time3, levels = design)

# Fit the new model
diff_species <- contrasts.fit(fit, cm1)
fit1 <- eBayes(diff_species)

# Pull the limma output for all genes

FDR_level <- 0.01

Time1v2 =topTable(fit1, coef=1, adjust="BH", number=Inf, sort.by="none")
Time2v3 =topTable(fit1, coef=2, adjust="BH", number=Inf, sort.by="none")

# Compare genes from T1-T2 to genes fromo T2-T3

plot(Time1v2$logFC, Time2v3$logFC)
plot(Time1v2$t, Time2v3$t)
plot(Time1v2$adj.P.Val, Time2v3$adj.P.Val)

# Look at the DE genes

dim(Time1v2[which(Time1v2$adj.P.Val < FDR_level),])
dim(Time2v3[which(Time2v3$adj.P.Val < FDR_level),])

simple_T1_T2 <- Time1v2[which(Time1v2$adj.P.Val < FDR_level),]
simple_T2_T3 <- Time2v3[which(Time2v3$adj.P.Val < FDR_level),]


# Gene to ID conversion document

gene_id <- read.table("../data/Homo_sapiens.GRCh38.v22_table.txt", stringsAsFactors = FALSE)
colnames(gene_id) <- c("ENSEMBL", "Gene_name")

# Eliminate the feature after the period (for DAVID)
check <- gsub("\\..*","",gene_id$ENSEMBL)

new_gene_id <- cbind(check, gene_id$Gene_name)
gene_id <- new_gene_id
colnames(gene_id) <- c("ENSEMBL", "Gene_name")

# Get gene names after filtering
genes=as.data.frame(rownames(Time1v2[which(Time1v2$adj.P.Val < FDR_level),]))
dim(genes)

colnames(genes) <- c("Gene_name")

# Get gene names of the list
comb_list <- merge(genes, gene_id, by = c("Gene_name"))
summary(duplicated(comb_list$Gene_name))

comb_list <- comb_list[!duplicated(comb_list$Gene_name),]

dim(comb_list)

write.table(comb_list$ENSEMBL, "../data/DAVID_2covar/simple_T1_T2_05.txt", quote = F, row.names = F, col.names = F)




# Get gene names after filtering
genes=as.data.frame(rownames(Time2v3[which(Time2v3$adj.P.Val < FDR_level),]))
dim(genes)

colnames(genes) <- c("Gene_name")

# Get gene names of the list
comb_list <- merge(genes, gene_id, by = c("Gene_name"))
summary(duplicated(comb_list$Gene_name))

comb_list <- comb_list[!duplicated(comb_list$Gene_name),]

dim(comb_list)

write.table(comb_list$ENSEMBL, "../data/DAVID_2covar/simple_T2_T3_05.txt", quote = F, row.names = F, col.names = F)



#write.table(simple_T1_T2$genes, "../data/DAVID_2covar/simple_T1_T2.txt")

head(topTable(fit1, coef=1, adjust="BH", number=100, sort.by="T"))
head(topTable(fit1, coef=2, adjust="BH", number=100, sort.by="T"))
```

## Make a Venn Diagram of the DE genes

```{r}
FDR_level <- 0.01

Time12 <- rownames(Time1v2[which(Time1v2$adj.P.Val < FDR_level),])

Time23 <- rownames(Time2v3[which(Time2v3$adj.P.Val < FDR_level),])

mylist <- list()
mylist[["DE T1 to T2"]] <- Time12
mylist[["DE T2 to T3"]] <- Time23

# Make as pdf 
Four_comp <- venn.diagram(mylist, filename= NULL, main="DE genes between timepoints (5% FDR)", cex=1.5 , fill = NULL, lty=1, height=2000, width=2000, rotation.degree = 0, scaled = FALSE, cat.pos = c(0,0))

grid.draw(Four_comp)
dev.off()


pdf(file = "~/Dropbox/Figures/DET1_T2.pdf")
  grid.draw(Four_comp)
dev.off()


```

# Re run using weight changes

```{r}
# Create the design matrix

# Use the standard treatment-contrasts parametrization. See Ch. 9 of limma
# User's Guide.

design <- model.matrix(~as.factor(Time) + as.factor(psychmeds) + RBC + RIN + Change_weight, data = clinical_sample)
colnames(design) <- c("Intercept", "Time2", "Time3", "Psychmeds", "RBC", "RIN", "Change_weight")

# Fit model 

# Model individual as a random effect.
# Recommended to run both voom and duplicateCorrelation twice.
# https://support.bioconductor.org/p/59700/#67620

cpm.voom <- voom(dge_in_cutoff, design, normalize.method="none")
#check_rel <- duplicateCorrelation(cpm.voom, design, block = clinical_sample$Individual)
check_rel_correlation <-  0.1687641
cpm.voom.corfit <- voom(dge_in_cutoff, design, normalize.method="none", plot = TRUE, block = clinical_sample$Individual, correlation = check_rel_correlation)
#check_rel <- duplicateCorrelation(cpm.voom.corfit, design, block = clinical_sample$Individual)
check_rel_correlation <- 0.1687548

# Look at the densities of the gene expression data 

plotDensities(cpm.voom.corfit[,1])
plotDensities(cpm.voom.corfit[,2])
plotDensities(cpm.voom.corfit[,3])
           
# Run lmFit and eBayes in limma

fit <- lmFit(cpm.voom.corfit, design, block=clinical_sample$Individual, correlation=check_rel_correlation)

# In the contrast matrix, have the time points

cm1 <- makeContrasts(Time1v2 = Time2, Time2v3 = Time3 - Time2, levels = design)

#cm1 <- makeContrasts(Time1v2 = Time2, Time2v3 = Time3, levels = design)

# Fit the new model
diff_species <- contrasts.fit(fit, cm1)
fit1 <- eBayes(diff_species)

# Pull the limma output for all genes

FDR_level <- 0.05

Time1v2 =topTable(fit1, coef=1, adjust="BH", number=Inf, sort.by="none")
Time2v3 =topTable(fit1, coef=2, adjust="BH", number=Inf, sort.by="none")

# Compare genes from T1-T2 to genes fromo T2-T3

plot(Time1v2$logFC, Time2v3$logFC)
plot(Time1v2$t, Time2v3$t)
plot(Time1v2$adj.P.Val, Time2v3$adj.P.Val)

# Look at the DE genes

dim(Time1v2[which(Time1v2$adj.P.Val < FDR_level),])
dim(Time2v3[which(Time2v3$adj.P.Val < FDR_level),])

head(topTable(fit1, coef=1, adjust="BH", number=100, sort.by="T"))
head(topTable(fit1, coef=2, adjust="BH", number=100, sort.by="T"))
```

## Make a Venn Diagram of the DE genes

```{r}
FDR_level <- 0.01

Time12 <- rownames(Time1v2[which(Time1v2$adj.P.Val < FDR_level),])

Time23 <- rownames(Time2v3[which(Time2v3$adj.P.Val < FDR_level),])

mylist <- list()
mylist[["DE T1 to T2"]] <- Time12
mylist[["DE T2 to T3"]] <- Time23

# Make as pdf 
Four_comp <- venn.diagram(mylist, filename= NULL, main="DE genes between timepoints (1% FDR)", cex=1.5 , fill = NULL, lty=1, height=2000, width=2000, rotation.degree = 180, scaled = FALSE, cat.pos = c(0,0))

grid.draw(Four_comp)
dev.off()


pdf(file = "~/Dropbox/Figures/DET1_T2_weight.pdf")
  grid.draw(Four_comp)
dev.off()


```



# Perform WGCNA on all DE genes (covariates not regressed out)

```{r}
# Load the package
library(WGCNA)
library(AnnotationDbi)
library(anRichment)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE)
enableWGCNAThreads()
allowWGCNAThreads()

# Pull the normalized gene expression for DE genes

genes12=as.data.frame(rownames(Time1v2[which(Time1v2$adj.P.Val < FDR_level),]))
colnames(genes12) <- c("DE_genes")

#genes23=as.data.frame(rownames(Time2v3[which(Time2v3$adj.P.Val < FDR_level),]))
#colnames(genes23) <- c("DE_genes")

de_genes <- genes12

check_gene <- rownames(cpm.voom.corfit$E) %in% de_genes$DE_genes
check_gene <- as.data.frame(check_gene)
pair_gene <- cbind(cpm.voom.corfit$E, check_gene)
norm_exp_T1T2 <- pair_gene[which(pair_gene$check_gene == TRUE),1:144]

dim(norm_exp_T1T2) #There are fewer rows than total DE genes because some genes are DE between T1 and T2 as well as T2 and T3

# Reshape the data so that we have 1 gene measurement per column and each sample name per row. 

transposed_norm_exp_T1T2 <- as.data.frame(t(norm_exp_T1T2))

dim(transposed_norm_exp_T1T2)
```

**Here, we have RNA seq data from 144 samples.** 

## Filter samples and genes with too many missing entries

This step is present in the WGCNA tutorial and is useful when using microarray data or unfiltered RNA seq data(which we are not). 

```{r}
# Rename so it matches
datExpr0 <- transposed_norm_exp_T1T2

gsg = goodSamplesGenes(datExpr0, verbose = 3);
gsg$allOK

if (!gsg$allOK)
{
  # Optionally, print the gene and sample names that were removed:
  if (sum(!gsg$goodGenes)>0) 
     printFlush(paste("Removing genes:", paste(names(datExpr0)[!gsg$goodGenes], collapse = ", ")));
  if (sum(!gsg$goodSamples)>0) 
     printFlush(paste("Removing samples:", paste(rownames(datExpr0)[!gsg$goodSamples], collapse = ", ")));
  # Remove the offending genes and samples from the data:
  datExpr0 = datExpr0[gsg$goodSamples, gsg$goodGenes]
}

```
**All of our samples passed this filtering step.**

# Sample clustering

This clustering is based on gene expression data for each sample. 

```{r}
sampleTree = hclust(dist(datExpr0), method = "average");
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)

# Plot a line to show the cut
abline(h = 15, col = "red")




# Determine cluster under the line

clust = cutreeStatic(sampleTree, cutHeight = 15, minSize = 1)
table(clust)
# clust 1 contains the samples we want to keep.
keepSamples = (clust==1)
datExpr = datExpr0[keepSamples, ]
nGenes = ncol(datExpr)
nSamples = nrow(datExpr)

# NOTE: In the WGCNA tutorial, cutHeight = 15. To start, we want to use all of the samples. Therefore, we will use the command below to maintain all samples

datExpr = datExpr0

```

## Add the clinical data (to match the WGCNA tutorial)

```{r}
datTraits = clinical_sample

datTraits[,1] <- as.numeric(datTraits[,1])
datTraits[,2] <- as.numeric(datTraits[,2])
datTraits[,4] <- as.numeric(datTraits[,4])
datTraits[,5] <- as.numeric(datTraits[,5])
datTraits[,6] <- as.numeric(datTraits[,6])

dim(datTraits)
```

## Cluster samples with gene expression and make heatmap of the clinical traits (to match the WGCNA tutorial)

```{r}
# Re-cluster samples
sampleTree2 = hclust(dist(datExpr), method = "average")
# Convert traits to a color representation: white means low, red means high, grey means missing entry
traitColors = numbers2colors(datTraits, signed = FALSE);
# Plot the sample dendrogram and the colors underneath.
plotDendroAndColors(sampleTree2, traitColors,
                    groupLabels = names(datTraits), 
                    main = "Sample dendrogram and trait heatmap")

save(datExpr, datTraits, file = "../data/FemaleWeightRestoration-01-dataInput.RData")
```

**Legend: white means low, red means high, grey means missing entry**

# Gene expression network construction and module detection

```{r}
# Load the data saved in the previous section
lnames = load(file = "../data/FemaleWeightRestoration-01-dataInput.RData")

# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to=20, by=2))
# Call the network topology analysis function
sft = pickSoftThreshold(datExpr, powerVector = powers, verbose = 5)
# Plot the results:
# Scale-free topology fit index as a function of the soft-thresholding power
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red")



# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity"))
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers,col="red")

softPower = 7;
adjacency = adjacency(datExpr, power = softPower)

# Turn adjacency into topological overlap
TOM = TOMsimilarity(adjacency);
dissTOM = 1-TOM
```

## Hierarchical clustering with TOM (Topological Overlap Matrix)-based dissimilarity

```{r}
geneTree = hclust(as.dist(dissTOM), method = "average");
# Plot the resulting clustering tree (dendrogram)
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04)

```

## Partion genes with similar gene expression into modules

```{r}
# We like large modules, so we set the minimum module size relatively high:
minModuleSize = 30;
# Module identification using dynamic tree cut:
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM, deepSplit = 2, pamRespectsDendro = FALSE, minClusterSize = minModuleSize);
table(dynamicMods)

# Give each module a color

# Convert numeric lables into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    main = "Gene dendrogram and module colors")

module_colors= setdiff(unique(dynamicColors), "grey")
for (color in module_colors){
    module=colnames(datExpr0)[which(dynamicColors==color)]
    
        # Convert to Ensembl gene ID
    
    # Get gene names after filtering
genes=as.data.frame(module)

colnames(genes) <- c("Gene_name")

# Get gene names of the list
comb_list <- merge(genes, gene_id, by = c("Gene_name"))
summary(duplicated(comb_list$Gene_name))

comb_list <- comb_list[!duplicated(comb_list$Gene_name),]
comb_list <- comb_list$ENSEMBL
    
    
#    write.table(comb_list, paste("../data/module_",color, ".txt",sep=""), sep="\t", row.names=FALSE, col.names=FALSE,quote=FALSE)
    
}
```

**Each color represents a different module.**


## Get the eigengenes for the project

```{r}
# Calculate eigengenes
MEList = moduleEigengenes(datExpr, colors = dynamicColors)
MEs = MEList$eigengenes

rownames(MEs) <- rownames(datExpr)

# Save the eigengenes

#write.table(MEs, #"../data/eigengenes_unadj_exp_9_modules.txt", quote = F)

# Quick look at the correlation between eigengenes and weight change

cor(MEs$MEblue, clinical_sample$Change_weight)
cor(MEs$MEbrown, clinical_sample$Change_weight)
cor(MEs$MEgreen, clinical_sample$Change_weight)
cor(MEs$MEgrey, clinical_sample$Change_weight)
cor(MEs$MEturquoise, clinical_sample$Change_weight)
cor(MEs$MEyellow, clinical_sample$Change_weight)

```

