---
title: "Info_by_timepoint"
author: "Lauren Blake"
date: "July 23, 2018"
output: workflowr::wflow_html
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

```{r cars}
# Load library for plotting
library(cowplot)
library(ggplot2)
library(gplots)

# Load data
ind_only <- read.csv("../data/clinical_sample_info.csv")
str(ind_only)
dim(ind_only)

summary(ind_only)
```

# Individuals per timepoint


```{r individuals per timepoint}
# T1
ind_only_t1 <- ind_only[which(ind_only$Time == 1),]

dim(ind_only_t1)

#T2
ind_only_t2 <- ind_only[which(ind_only$Time == 2),]

dim(ind_only_t2)

#T3
ind_only_t3 <- ind_only[which(ind_only$Time == 3),]

dim(ind_only_t3)

summary(as.factor(ind_only_t3$current_ED))

#T4
ind_only_t4 <- ind_only[which(ind_only$Time == 4),]

dim(ind_only_t4)

summary(as.factor(ind_only_t4$current_ED))

#T5
ind_only_t5 <- ind_only[which(ind_only$Time == 5),]

dim(ind_only_t5)

summary(as.factor(ind_only_t5$current_ED))
```

# Race/ethnicity at T1

```{r}
summary(as.factor(ind_only_t1$Race))
summary(as.factor(ind_only_t1$Ethnicity))

```

# Admission characteristics

```{r}
no_time <- read.csv("../data/notimecovariates.csv")
str(no_time)

# Age of onset

summary(no_time$age_onset)
mean(no_time$age_onset[-47])
sd(no_time$age_onset[-47])

# Age

summary(ind_only_t1$age)
mean(ind_only_t1$age)
sd(ind_only_t1$age)

# Prior hospitalizations

prev_hosp <- no_time[which(no_time$Previous_hospitalizations != "NA"),]

summary(prev_hosp$Previous_hospitalizations)
mean(prev_hosp$Previous_hospitalizations)
sd(prev_hosp$Previous_hospitalizations)

# Subtype

summary(as.factor(no_time$AN_subtype))

# Weight

summary(ind_only_t1$Weight)
mean(ind_only_t1$Weight)
sd(ind_only_t1$Weight)

# BMI

summary(ind_only_t1$bmi)
mean(ind_only_t1$bmi)
sd(ind_only_t1$bmi)


# EDU
summary(no_time$EDU_days)
mean(no_time$EDU_days)
sd(no_time$EDU_days)

# AMA
summary(as.factor(no_time$AMA))

# AMA versus EDU days

summary(lm(no_time$EDU_days ~ as.factor(no_time$AMA)))

# Weight over time

summary(ind_only_t1$Weight)
summary(ind_only_t2$Weight)
summary(ind_only_t3$Weight)
summary(ind_only_t4$Weight)
summary(ind_only_t5$Weight)
```

# Change in BMI

```{r}
#T1 to T2

summary(no_time$bmiT1T2)
mean(no_time$bmiT1T2)
sd(no_time$bmiT1T2)

#T2 to T3

time <- no_time[which(no_time$bmiT2T3 != "NA"),]

summary(time$bmiT2T3)
mean(time$bmiT2T3)
sd(time$bmiT2T3)

#T3 to T4

time <- no_time[which(no_time$bmiT3T4 != "NA"),]

summary(time$bmiT3T4)
mean(time$bmiT3T4)
sd(time$bmiT3T4)

#T4 to T5

time <- no_time[which(no_time$bmiT4T5 != "NA"),]

summary(time$bmiT4T5)
mean(time$bmiT4T5)
sd(time$bmiT4T5)
```

# Correlation between days on unit and days between blood draws

```{r}
cor(no_time$EDU_days, no_time$blooddrawT1T2)
```

# Correlation between days on unit and bmiT1T2

```{r}
cor(no_time$EDU_days, no_time$bmiT1T2)
```

# Correlation between days between T1T2 blood draws and bmiT1T2

```{r}
cor(no_time$blooddrawT1T2, no_time$bmiT1T2)

weight_over_time <- ggplot(no_time, aes(no_time$EDU_days, no_time$bmiT1T2)) + geom_point() + geom_smooth(method='lm',formula=y~x)  + theme(legend.position = "none") + xlab("EDU days") + ylab("Change in BMI (T1 to T2)") 

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/EDU_over_BMI_boxplot.png", weight_over_time,
#          base_aspect_ratio = 1)

weight_over_time <- ggplot(no_time, aes(no_time$EDU_days, no_time$bmiT1T2)) + geom_point()   + theme(legend.position = "none") + xlab("EDU days") + ylab("Change in BMI (T1 to T2)") 

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/EDU_over_BMI_boxplot_no_line.png", weight_over_time,
#          base_aspect_ratio = 1)


# AMA status
weight_over_time <- ggplot(no_time, aes(as.factor(no_time$AMA), no_time$bmiT1T2)) + geom_boxplot(outlier.shape = NA) + geom_jitter(aes(color = as.factor(no_time$AMA)), width = 0.1) + xlab("AMA") + ylab("Change in BMI (T1 to T2)") + theme(legend.position = "none") + scale_x_discrete(breaks=c("0","1"),
        labels=c("No", "Yes"))

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/AMA_BMI_T1T2boxplot.png", weight_over_time,
#          base_aspect_ratio = 1)


summary(lm(no_time$bmiT1T2 ~ as.factor(no_time$AMA)))

weight_over_time <- ggplot(no_time, aes(as.factor(no_time$AMA), no_time$bmiT2T3)) + geom_boxplot(outlier.shape = NA) + geom_jitter(aes(color = as.factor(no_time$AMA)), width = 0.1) + xlab("AMA") + ylab("Change in BMI (T2 to T3)") + theme(legend.position = "none") + scale_x_discrete(breaks=c("0","1"),
        labels=c("No", "Yes"))

plot_grid(weight_over_time)

#save_plot("/Users/laurenblake/Dropbox/Lauren Blake/Figures/AMA_BMI_T2T3boxplot.png", weight_over_time,
#          base_aspect_ratio = 1)

summary(lm(no_time$bmiT2T3 ~ as.factor(no_time$AMA)))



```

# Correlation between lab results

```{r}
ind_only_values <- c(12, 13, 21, 22, 23, 24, 25, 51)

ind_only_labs <- ind_only[,ind_only_values]

ind_only_labs <- as.data.frame(ind_only_labs)
  
# All labs together

  # Platelet 

cor(ind_only_labs$Platelet, ind_only_labs$RBC, use = "pairwise.complete.obs")

cor(ind_only_labs$Platelet, ind_only_labs$AN, use = "pairwise.complete.obs")

cor(ind_only_labs$Platelet, ind_only_labs$AL, use =  "pairwise.complete.obs")

cor(ind_only_labs$Platelet, ind_only_labs$AE, use = "pairwise.complete.obs")

  # RBC

cor(ind_only_labs$RBC, ind_only_labs$AN, use = "pairwise.complete.obs")

cor(ind_only_labs$RBC, ind_only_labs$AL, use = "pairwise.complete.obs")

cor(ind_only_labs$RBC, ind_only_labs$AE, use =  "pairwise.complete.obs")

  # AN

cor(ind_only_labs$AN, ind_only_labs$AL, use =  "pairwise.complete.obs")

cor(ind_only_labs$AN, ind_only_labs$AE, use =  "pairwise.complete.obs")

  # AL

cor(ind_only_labs$AL, ind_only_labs$AE, use =  "pairwise.complete.obs")



check_cor_labs <- ind_only_labs


pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

#Find which correlations are < 

which(abs(correlations) > 0.5)

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))

#png("/Users/laurenblake/Dropbox/Lauren Blake/Figures/5_values_all_times.png",
#  width = 5*300,        # 5 x 300 pixels
#  height = 5*300,
#  res = 300,            # 300 pixels per inch
#  pointsize = 8)  

heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none", main = "Correlations between \n lab values, all time points", notecex = 2, trace = "none", key = TRUE, margins = c(12,8))

dev.off()
```

# Separate and look at the correlations bet. lab values at each point

```{r}
# Time point 1

ind_only_values <- c(2,3,12, 13, 21, 22, 23, 24, 25, 51)

ind_only_labs <- ind_only[,ind_only_values]

check_cor_labs <- ind_only_labs[which(ind_only_labs$Time == 1), -1]

check_cor_labs <- check_cor_labs[,-1]

pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))

#png("/Users/laurenblake/Dropbox/Lauren Blake/Figures/5_values_T1.png",
#  width = 5*300,        # 5 x 300 pixels
#  height = 5*300,
#  res = 300,            # 300 pixels per inch
#  pointsize = 8)  

heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none", main = "Correlations between \n lab values, T1", notecex = 2, trace = "none", key = TRUE, margins = c(12,8))

dev.off()




# Time point 2

ind_only_values <- c(2, 3, 12, 13, 21, 22, 23, 24, 25, 51)

ind_only_labs <- ind_only[,ind_only_values]

check_cor_labs <- ind_only_labs[which(ind_only_labs$Time == 2), -1]

check_cor_labs <- check_cor_labs[,-1]

pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))

#png("/Users/laurenblake/Dropbox/Lauren Blake/Figures/5_values_T2.png",
#  width = 5*300,        # 5 x 300 pixels
#  height = 5*300,
#  res = 300,            # 300 pixels per inch
#  pointsize = 8)  

heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none", main = "Correlations between \n lab values, T2", notecex = 2, trace = "none", key = TRUE, margins = c(12,8))

dev.off()


# Time point 3

ind_only_values <- c(2, 3, 12, 20, 21, 22, 23, 24, 51)

ind_only_labs <- ind_only[,ind_only_values]

check_cor_labs <- ind_only_labs[which(ind_only_labs$Time == 3), -1]

check_cor_labs <- check_cor_labs[,-1]

pvalues <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(pvalues) <- colnames(check_cor_labs)
rownames(pvalues) <- colnames(check_cor_labs)
correlations <- matrix(data = NA, nrow = ncol(check_cor_labs), ncol = ncol(check_cor_labs))
colnames(correlations) <- colnames(check_cor_labs)
rownames(correlations) <- colnames(check_cor_labs)

j=1
for (j in 1:ncol(check_cor_labs)){
  
  for (i in 1:ncol(check_cor_labs)){
      
  test <- cor.test(as.numeric(check_cor_labs[,j]), as.numeric(check_cor_labs[,i]), method = "pearson")
      
  #Get the correlation
  
  test$estimate
  
  #Get the p-value 
  test$p.value

  pvalues[j, i] <- test$p.value
  correlations[j, i] <- test$estimate
  
  i = i+1
  }
  j=j+1
}

#correlations
#pvalues

#Find which variables are p-value < 0.05
pvalues <=0.05

# need data as matrix
mm <- as.matrix(correlations, ncol = ncol(correlations))

#png("/Users/laurenblake/Dropbox/Lauren Blake/Figures/5_values_T3.png",
#  width = 5*300,        # 5 x 300 pixels
#  height = 5*300,
#  res = 300,            # 300 pixels per inch
#  pointsize = 8)  

heatmap.2(x = mm, Rowv = FALSE, Colv = FALSE, dendrogram = "none", main = "Correlations between \n lab values, T3", notecex = 2, trace = "none", key = TRUE, margins = c(12,8))

dev.off()

```


# How much do the individual lab values change over time?

```{r}
# Include individual
ind_only_values <- c(2,3,12, 13, 21, 22, 23, 24, 25, 51)

ind_only_labs <- ind_only[,ind_only_values]

# Stratify the 5 lab values by time point
check_cor_labs1 <- ind_only_labs[which(ind_only_labs$Time == 1), ]

check_cor_labs2 <- ind_only_labs[which(ind_only_labs$Time == 2), ]

# Time points 1 and 2

cor(check_cor_labs1$RBC, check_cor_labs2$RBC, "pairwise.complete.obs")

cor(check_cor_labs1$Platelet, check_cor_labs2$Platelet, "pairwise.complete.obs")

cor(check_cor_labs1$AN, check_cor_labs2$AN, "pairwise.complete.obs")

cor(check_cor_labs1$AE, check_cor_labs2$AE, "pairwise.complete.obs")

cor(check_cor_labs1$AL, check_cor_labs2$AL, "pairwise.complete.obs")

cor(check_cor_labs1$AB, check_cor_labs2$AB, "pairwise.complete.obs")

cor(check_cor_labs1$AM, check_cor_labs2$AM, "pairwise.complete.obs")

# Time points 2 and 3

check_cor_labs2 <- ind_only_labs[which(ind_only_labs$Time == 2), ]
check_cor_labs3 <- ind_only_labs[which(ind_only_labs$Time == 3), ]

check_cor_labs23 <- merge(check_cor_labs2, check_cor_labs3, by = "BAN_ID")

cor(check_cor_labs23$RBC.x, check_cor_labs23$RBC.y, "pairwise.complete.obs")

cor(check_cor_labs23$Platelet.x, check_cor_labs23$Platelet.y, "pairwise.complete.obs")

cor(check_cor_labs23$AN.x, check_cor_labs23$AN.y, "pairwise.complete.obs")

cor(check_cor_labs23$AE.x, check_cor_labs23$AE.y, "pairwise.complete.obs")

cor(check_cor_labs23$AL.x, check_cor_labs23$AL.y, "pairwise.complete.obs")

cor(check_cor_labs23$AB.x, check_cor_labs23$AB.y, "pairwise.complete.obs")

cor(check_cor_labs23$AM.x, check_cor_labs23$AM.y, "pairwise.complete.obs")


# T0 to T1
# Import T0
check_cor_labs0 <- read.csv("../data/T0_consolid.csv", stringsAsFactors = FALSE)

check_cor_labs1 <- ind_only_labs[which(ind_only_labs$Time == 1), ]

# Merge T0 and T1

cor(check_cor_labs1$RBC, as.numeric(check_cor_labs0$RBC), "pairwise.complete.obs")

cor(check_cor_labs1$Platelet, check_cor_labs0$Platelet, "pairwise.complete.obs")

cor(check_cor_labs1$AN, check_cor_labs0$AN, "pairwise.complete.obs")

cor(check_cor_labs1$AE, check_cor_labs0$AE, "pairwise.complete.obs")

cor(check_cor_labs1$AL, check_cor_labs0$AL, "pairwise.complete.obs")

cor(check_cor_labs1$AB, check_cor_labs0$AB, "pairwise.complete.obs")

cor(check_cor_labs1$AM, check_cor_labs0$AM, "pairwise.complete.obs")

```

